A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-01-18 paper/paper.tex
More progress
progress

\title{A formalization of semi-simplicial and semi-cubical sets in dependent type theory}
\author{\IEEEauthorblockN{Hugo Herbelin \\ and Ramkumar Ramachandra} \IEEEauthorblockA{Université Paris Cité, Inria, CNRS, IRIF, Paris}}

\begin{document}
\maketitle
\begin{abstract}
  Constructions such as semi-simplicial and semi-cubical sets can be defined in the ``usual way'' as presheaves over respectively the semi-simplex or semi-cube category, which we shall call \emph{fibered} definitions, but also, as done e.g. in Voevodsky~\cite{voevodsky12} and in previous work~\cite{herbelin15} as a dependently-typed, or, say, \emph{indexed} construction.

  This paper describes an alternative indexed construction of both augmented semi-simplicial and semi-cubical sets based on the property that they naturally arise as respectively unary and binary iterated parametricity-based constructions.

  Additionally, the axiom-free construction has been fully formalized in Coq's dependent type theory, which turned out not to be so straightforward.
\end{abstract}

@@

Our precise mechanization in Coq is available on Github, at \href{https://github.com/artagnon/bonak}{github.com/artagnon/bonak}.

@@

@@

>>>>

\title{A parametricity-based formalization of semi-simplicial and semi-cubical sets}
\author{\IEEEauthorblockN{Hugo Herbelin \\ and Ramkumar Ramachandra} \IEEEauthorblockA{Université Paris Cité, Inria, CNRS, IRIF, Paris}}

\begin{document}
\maketitle
\begin{abstract}
  Constructions such as semi-simplicial and semi-cubical sets can be defined in the ``usual way'' as presheaves over respectively, the semi-simplex or semi-cube category, which we call \emph{fibered} definitions, but also defined like in e.g. Voevodsky~\cite{voevodsky12} or in previous work~\cite{herbelin15}, as a dependently-typed construction, which we call \emph{indexed}.

  This paper describes an alternative indexed characterization of both augmented semi-simplicial and semi-cubical sets arising respectively as unary and binary iterated parametricity-based constructions.

  Additionally, our construction is fully formalized in Coq's dependent type theory.
\end{abstract}

@@

Our precise mechanization in Coq is available on Github, at \href{https://github.com/artagnon/bonak}{github.com/artagnon/bonak}. The construction was conceived in Summer 2019 and the mechanization began in late 2019. A sketch of the construction was presented at the 2020 HoTT-UF workshop and the completion of the mechanization was reported at the TYPES 2022 conference.

@@

Now, let's observe that the process of construction of the type of
$X_1$ from that of $X_0$ and from that of $X_2$ to that of $X_1$ is
similar to applying a binary parametricity translation on contexts that
is: to triplicate each declaration $x_L:X_0$ and $x_R:X_0$ into two
copies of $x$ and a line connecting these copies, that is into two
triples of declarations $x_{LL}:X_0, x_{LR}:X_0,
x_{L\kstar}:X_1(x_{LL},x_{LR})$ and $x_{RL}:X_0, x_{RL}:X_0,
x_{R\kstar}:X_1(x_{RL},x_{RR})$, and to expand $\U$ into the type of a
relation over these two copies, that is, into $x_{\kstar
  L}:X_1(x_{LL},x_{RL}) \times x_{\kstar R}:X_1(x_{LR},x_{RR})
\rightarrow \U$.

@@

\section{Perspectives}

In the cubical case, we expect the construction to eventually provide
a model of (some version of) parametric type theory~\cite{} by adding
degeneracies, a hierarchy of universes (as sketched e.g. in a talk by
Herbelin at the HoTT-UF workshop for the bridge
case~\cite{herbelin-hott-uf}) as well as reasoning modulo
permutations~\cite{grandis03}.

By equipping the universe construction with a structure of
equivalences (as suggested along the lines of Altenkirch and
Kaposi~\cite{}), as well as connections~\cite{}, we expect the
parametric model to extend into a model of cubical type
theory~\cite{}. A key point in this approach, is that, thanks to the
indexed construction, we expect univalence to hold by definition in
the model, without requiring any kind of Reedy fibrancy as it would a
priori be needed in the fibered case. If these intuitions are correct,
this would definitively root cubical type theory on top of iterated
parametricity translation.
